---
title: "Exploratory and Inferential Analysis of Compliance Behaviour Across Nigerian Tax Obligations"
author: "Lilian Obadan"
date: today
format:
  html:
    theme: flatly
    toc: true
    code-fold: true
    self-contained: true
---

```{=html}
<style>
:root {
  --navy:       #1a3a5c;
  --navy-dark:  #0d2b4e;
  --navy-mid:   #1e5080;
  --navy-light: #eaf1f8;
  --gold:       #f0a500;
}
body { font-family: Calibri, sans-serif; color: #1c2340; background-color: #f7f9fc; line-height: 1.7; }
.quarto-title-block {
  background: linear-gradient(135deg, #0d2b4e 0%, #1a3a5c 55%, #1e5080 100%);
  color: white; padding: 2.5rem 3rem; border-radius: 0 0 16px 16px; margin-bottom: 2rem;
}
.quarto-title { color: white !important; font-size: 1.6rem; }
.quarto-title-meta { color: #a8c6e8 !important; }
#TOC { background-color: #1a3a5c; padding: 14px 16px; border-radius: 10px; color: white; font-size: 13px; }
#TOC a { color: #a8c8e8 !important; text-decoration: none; }
#TOC a:hover { color: #f0a500 !important; }
.toc-title { color: #f0a500 !important; font-weight: bold; text-transform: uppercase; letter-spacing: 1px; }
h1 { color: #1a3a5c; font-size: 1.45rem; }
h2 { color: #1a3a5c; border-left: 5px solid #f0a500; padding-left: 10px; margin-top: 2rem; font-size: 1.2rem; }
h3 { color: #1e5080; font-size: 1.05rem; margin-top: 1.4rem; }
thead tr { background-color: #1a3a5c !important; color: white !important; }
tbody tr:nth-child(even) { background-color: #eaf1f8; }
.callout-note    { border-left-color: #1a3a5c !important; }
.callout-tip     { border-left-color: #1b5e20 !important; }
.callout-warning { border-left-color: #f0a500 !important; }
.callout-important { border-left-color: #b71c1c !important; }
pre { background-color: #f0f5fb !important; border-left: 4px solid #1a3a5c; border-radius: 6px; }
hr { border: none; height: 2px; background: linear-gradient(to right, #f0a500, transparent); margin: 2rem 0; }
</style>
```

```{r setup, include=FALSE}
# ── Reproducibility ────────────────────────────────────────────────────────────
set.seed(2025)
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", out.width = "100%", fig.height = 5
)

# ── Packages ───────────────────────────────────────────────────────────────────
library(tidyverse)
library(skimr)
library(janitor)
library(scales)
library(ggcorrplot)
library(broom)
library(kableExtra)
library(car)
library(lmtest)
library(rstatix)

# ── CRITICAL: Fix namespace conflicts ──────────────────────────────────────────
# car loads MASS, which masks dplyr::select; rstatix masks dplyr::filter.
# These two lines force dplyr's versions to win.
select <- dplyr::select
filter <- dplyr::filter

# ── Colours ────────────────────────────────────────────────────────────────────
NAVY  <- "#1a3a5c"
GOLD  <- "#f0a500"
LIGHT <- "#eaf1f8"
RED   <- "#b71c1c"

pal_tax  <- c("VAT" = "#1a5276", "PAYE" = "#2e86c1",
              "WHT" = "#f39c12", "CIT"  = "#c0392b")
pal_comp <- c("On Time" = "#1b5e20", "Delayed" = "#c0392b")
pal_band <- c("Band 1 < N5B" = "#76b7b2", "Band 2 N5-15B" = "#4e79a7",
              "Band 3 N15-35B" = "#f28e2b", "Band 4 > N35B" = "#e15759")
pal_mgmt <- c("Local" = "#1a3a5c", "Foreign" = "#f0a500")

# ── ggplot2 theme ──────────────────────────────────────────────────────────────
theme_lbs <- function(base_size = 12) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.title       = element_text(face = "bold", colour = NAVY, size = base_size + 1, margin = margin(b = 5)),
      plot.subtitle    = element_text(colour = "#5a6282", size = base_size - 1, margin = margin(b = 8)),
      axis.title       = element_text(face = "bold", colour = NAVY, size = base_size - 1),
      axis.text        = element_text(colour = "#5a6282", size = base_size - 2),
      panel.background = element_rect(fill = "#f7f9fc", colour = NA),
      plot.background  = element_rect(fill = "#f7f9fc", colour = NA),
      strip.text       = element_text(face = "bold", colour = NAVY),
      strip.background = element_rect(fill = LIGHT, colour = NA),
      legend.position  = "bottom"
    )
}
theme_set(theme_lbs())
```

```{r data-load, include=FALSE}
# ── Load and clean the dataset ─────────────────────────────────────────────────
# The CSV must be in the same folder as this .qmd file.
df_raw <- read.csv("DA_Exam_data_for_upload_new.csv",
                   stringsAsFactors = FALSE,
                   strip.white      = TRUE)

df <- df_raw |>
  clean_names() |>
  mutate(
    turnover_ngn  = as.numeric(gsub(",", "", company_turnover)),
    amount_paid_n = suppressWarnings(as.numeric(gsub(",", "", amount_paid))),
    date_paid     = as.Date(date_paid, format = "%d/%m/%Y"),
    filing_period = as.factor(filing_period),
    revenue_band  = cut(turnover_ngn / 1e9,
                        breaks = c(0, 5, 15, 35, Inf),
                        labels = c("Band 1 < N5B", "Band 2 N5-15B",
                                   "Band 3 N15-35B", "Band 4 > N35B"),
                        include.lowest = TRUE),
    compliance_status = factor(
      if_else(filing_delay_days == 0, "On Time", "Delayed"),
      levels = c("On Time", "Delayed")),
    tax_type        = factor(tax_type, levels = c("VAT","PAYE","WHT","CIT")),
    management_type = factor(management_type, levels = c("Local","Foreign")),
    industry_sector = factor(industry_sector)
  )
```

---

## 1. Executive Summary

This study examines filing compliance behaviour for 104 tax filing observations across 13
companies with major Nigerian tax obligations including Value Added Tax (VAT),
Pay-As-You-Earn (PAYE), Withholding Tax (WHT), and Companies Income Tax (CIT). The
business problem is that companies operating within the same regulatory environment may
demonstrate inconsistent filing behaviour across different tax obligations, creating
delayed remittances and reconciliation challenges for the companies and their tax
consultants, as well as compliance inefficiencies and potential revenue leakage to the
tax authorities.

Using anonymised operational tax compliance data collected from tax returns filed on
behalf of clients, this study applies five analytical techniques: Exploratory Data
Analysis, Data Visualisation, Hypothesis Testing, Correlation Analysis, and Regression
Analysis. The purpose is to understand whether tax compliance — measured by filing delay
behaviour — differs across tax types and whether other operational variables such as
filing frequency, company turnover, amount paid, industry sector, and management type are
associated with compliance behaviour.

The study finds that compliance varies significantly across tax types. VAT achieves the
highest level of compliance with a 100% on-time filing rate, attributable to its clear
monthly filing deadlines, automatic penalties for late filing, and the high frequency of
filing which builds habitual compliance behaviour. CIT records the highest average delay
and the most severe individual non-compliance event (677 days), attributable to the
substantial cash outlay required, the infrequent annual filing cycle, and the complexity
of tax computation. A major recommendation arising from this study is the introduction of
monthly filing of estimated CIT payments by all companies, similar to the practice already
established in the upstream oil and gas sector.

---

## 2. Professional Disclosure

I work as a tax compliance professional within the tax advisory and compliance sector,
where I am responsible for monitoring filing timelines, reviewing statutory obligations,
supporting taxpayers, analysing compliance behaviour, and assisting companies in meeting
their compliance goals. These responsibilities are directly relevant to my day-to-day
work and form the operational context from which this dataset was drawn. This study is
directly connected to my professional practice because it uses operational tax filing
data to examine compliance behaviour across VAT, PAYE, WHT, and CIT obligations.

**Exploratory Data Analysis** is relevant because tax compliance data typically contains
filing variations, outliers, and distributional patterns that must first be understood
before any deeper analysis is conducted.

**Data Visualisation** is relevant because tax findings must frequently be communicated
clearly to clients, management teams, and regulators who are not statisticians.

**Hypothesis Testing** is relevant because it allows me to determine whether observed
differences in filing delays across tax types are statistically meaningful.

**Correlation Analysis** is relevant because filing delay behaviour may be associated
with operational variables such as filing frequency, turnover, and amount paid.

**Regression Analysis** is relevant because it estimates the independent contribution of
each factor to filing delay, enabling targeted, prioritised recommendations.

---

## 3. Data Collection & Sampling

The dataset was collected from anonymised operational tax compliance records relating to
13 selected companies and their filing behaviour across VAT, PAYE, WHT, and CIT
obligations. Each observation represents a unique company–tax type–filing period
combination, yielding 104 records across four filing periods from January 2024 to July 2025.

The sampling method is **purposive sampling**: data was extracted from compliance tracking
registers maintained in the normal course of professional tax advisory work. All company
identifiers were anonymised and replaced with codes (Company A through Company M) before
analysis. No taxpayer identification numbers, personal identifiers, or confidential client
details are disclosed.

```{r data-shape}
cat("Rows:", nrow(df), "| Columns:", ncol(df), "\n")
cat("Missing values per column:\n")
print(colSums(is.na(df)))
```

The 104 observations exceed the assessment minimum of 100 and support all five inferential
techniques applied. The dataset is a census of available filing records for the 13 clients
under active engagement during the observation period, which maximises completeness but
limits generalisation to the broader Nigerian corporate taxpayer population.

---

## 4. Data Description

### Variable Dictionary

```{r var-dict, echo=FALSE}
tibble(
  Variable    = c("Client","Filing_Delay_Days","Filing_Period","Tax_Type",
                  "Filing_Frequency","Industry_Sector","Company_Turnover",
                  "Amount_Paid","Date_Paid","Management_Type",
                  "compliance_status","revenue_band"),
  Type        = c("Categorical","Numeric","Categorical","Categorical",
                  "Numeric","Categorical","Numeric","Numeric",
                  "Date","Categorical","Derived","Derived"),
  Description = c(
    "Anonymised company code (A-M); 13 unique companies",
    "Days between statutory due date and actual filing; 0 = on time",
    "Tax filing period; 4 periods covered",
    "Statutory obligation: VAT, PAYE, WHT, or CIT",
    "Required filings per year: 12 monthly | 1 annual (CIT)",
    "Industry: ICT or Oil and Gas",
    "Annual turnover in Naira",
    "Amount remitted in Naira; 20 entries are nil",
    "Date payment was received",
    "Local or Foreign management",
    "On Time vs Delayed (binary outcome)",
    "Four turnover tiers"
  )
) |>
  kbl(booktabs = TRUE, caption = "Table 1 — Variable Dictionary") |>
  kable_styling(full_width = TRUE, bootstrap_options = c("striped","hover")) |>
  row_spec(0, bold = TRUE, background = NAVY, color = "white") |>
  column_spec(1, bold = TRUE)
```

### Distributional Summary

```{r skim-output}
df |>
  dplyr::select(filing_delay_days, turnover_ngn, amount_paid_n,
                filing_frequency, tax_type, industry_sector,
                management_type, compliance_status, revenue_band) |>
  skim()
```

### Observations by Tax Type

```{r tab-tax, echo=FALSE}
df |>
  dplyr::count(tax_type, name = "n") |>
  dplyr::mutate(`%` = scales::percent(n / sum(n), accuracy = 0.1)) |>
  dplyr::rename(`Tax Type` = tax_type) |>
  kbl(booktabs = TRUE, caption = "Table 2a — Observations by Tax Type") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover")) |>
  row_spec(0, bold = TRUE, background = NAVY, color = "white")
```

### Observations by Industry Sector

```{r tab-ind, echo=FALSE}
df |>
  dplyr::count(industry_sector, name = "n") |>
  dplyr::mutate(`%` = scales::percent(n / sum(n), accuracy = 0.1)) |>
  dplyr::rename(`Industry` = industry_sector) |>
  kbl(booktabs = TRUE, caption = "Table 2b — Observations by Industry Sector") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover")) |>
  row_spec(0, bold = TRUE, background = NAVY, color = "white")
```

### Observations by Management Type

```{r tab-mgmt, echo=FALSE}
df |>
  dplyr::count(management_type, name = "n") |>
  dplyr::mutate(`%` = scales::percent(n / sum(n), accuracy = 0.1)) |>
  dplyr::rename(`Management Type` = management_type) |>
  kbl(booktabs = TRUE, caption = "Table 2c — Observations by Management Type") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover")) |>
  row_spec(0, bold = TRUE, background = NAVY, color = "white")
```

### Observations by Revenue Band

```{r tab-band, echo=FALSE}
df |>
  dplyr::count(revenue_band, name = "n") |>
  dplyr::mutate(`%` = scales::percent(n / sum(n), accuracy = 0.1)) |>
  dplyr::rename(`Revenue Band` = revenue_band) |>
  kbl(booktabs = TRUE, caption = "Table 3 — Observations by Revenue Band") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover")) |>
  row_spec(0, bold = TRUE, background = NAVY, color = "white")
```

---

## 5. Exploratory Data Analysis

> **Technique 1 of 5** — Summary statistics, missing-value analysis, outlier detection

::: {.callout-note title="Business Justification"}
Before any compliance recommendation can be made, the data must be examined for quality
issues, distributional irregularities, and anomalies. This step mirrors the mandatory
pre-engagement data review I conduct before submitting client representations to FIRS.
:::

### 5.1 Summary Statistics

```{r eda-summary, echo=FALSE}
delay_vec <- df$filing_delay_days

tibble(
  Statistic = c("N", "Mean", "Median", "Std Dev", "Min",
                "Q1 (25%)", "Q3 (75%)", "Max",
                "% On Time (delay = 0)", "Skewness"),
  Value = c(
    as.character(length(delay_vec)),
    format(round(mean(delay_vec), 2), nsmall = 2),
    as.character(median(delay_vec)),
    format(round(sd(delay_vec), 2), nsmall = 2),
    as.character(min(delay_vec)),
    as.character(quantile(delay_vec, 0.25)),
    as.character(quantile(delay_vec, 0.75)),
    as.character(max(delay_vec)),
    paste0(round(mean(delay_vec == 0) * 100, 1), "%"),
    format(round(
      (sum((delay_vec - mean(delay_vec))^3) / length(delay_vec)) /
        sd(delay_vec)^3, 2), nsmall = 2)
  )
) |>
  kbl(caption = "Table 4 — Descriptive Statistics: Filing_Delay_Days",
      booktabs = TRUE) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover")) |>
  row_spec(0, bold = TRUE, background = NAVY, color = "white") |>
  row_spec(c(2, 9), bold = TRUE)
```

### 5.2 Filing Delay by Tax Type

```{r eda-by-type, echo=FALSE}
df |>
  dplyr::group_by(`Tax Type` = tax_type) |>
  dplyr::summarise(
    n              = dplyr::n(),
    `Mean Delay`   = round(mean(filing_delay_days), 2),
    `Median Delay` = median(filing_delay_days),
    `Min`          = min(filing_delay_days),
    `Max`          = max(filing_delay_days),
    `% On Time`    = paste0(round(mean(filing_delay_days == 0) * 100, 1), "%"),
    .groups        = "drop"
  ) |>
  kbl(caption = "Table 5 — Filing Delay Summary by Tax Type", booktabs = TRUE) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover")) |>
  row_spec(0, bold = TRUE, background = NAVY, color = "white")
```

### 5.3 Outlier Detection

```{r outliers, echo=FALSE}
iqr_val   <- IQR(df$filing_delay_days)
hi_fence  <- quantile(df$filing_delay_days, 0.75) + 1.5 * iqr_val

df_out <- df |>
  dplyr::filter(filing_delay_days > hi_fence) |>
  dplyr::select(client, tax_type, filing_period,
                filing_delay_days, revenue_band, management_type) |>
  dplyr::arrange(dplyr::desc(filing_delay_days))

cat("IQR upper fence:", hi_fence, "| Outliers flagged:", nrow(df_out), "\n\n")

df_out |>
  kbl(caption = "Table 6 — Outlier Observations (IQR Method)", booktabs = TRUE) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover")) |>
  row_spec(0, bold = TRUE, background = NAVY, color = "white")
```

::: {.callout-warning title="Data Quality Issues"}
**Issue 1 — Nil Amount_Paid entries (20 records, 19.2%):** filings with zero remittance,
representing nil-return obligations. Treated as valid zero values, not missing data.

**Issue 2 — Right-skewed, zero-inflated outcome:** 75% of filings carry no delay; the
mean of 21.3 days is distorted by extreme CIT outliers.

**Issue 3 — Extreme outlier at 677 days:** One CIT filing — a genuine compliance failure,
not a data error. Retained in the analysis with appropriate acknowledgement.
:::

---

## 6. Data Visualisation

> **Technique 2 of 5** — Grammar of graphics, chart selection, storytelling

### Figure 1 — Distribution of Filing Delay Days

```{r fig1, echo=FALSE, fig.height=4.5}
ggplot(df, aes(x = filing_delay_days)) +
  geom_histogram(aes(y = after_stat(density)), bins = 28,
                 fill = NAVY, colour = "white", alpha = 0.85) +
  geom_density(colour = GOLD, linewidth = 1.2) +
  scale_x_continuous(labels = scales::comma) +
  labs(title    = "Figure 1 — Distribution of Filing Delay Days",
       subtitle = "Zero-inflated and right-skewed; 75% of filings carry no delay",
       x = "Filing Delay (Days)", y = "Density")
```

### Figure 2 — Compliance Rate by Tax Type

```{r fig2, echo=FALSE, fig.height=4.5}
df |>
  dplyr::count(tax_type, compliance_status) |>
  dplyr::group_by(tax_type) |>
  dplyr::mutate(pct = n / sum(n)) |>
  ggplot(aes(x = tax_type, y = pct, fill = compliance_status)) +
  geom_col(width = 0.6, colour = "white") +
  geom_text(aes(label = scales::percent(pct, accuracy = 1)),
            position = position_stack(vjust = 0.5),
            colour = "white", size = 3.8, fontface = "bold") +
  scale_fill_manual(values = pal_comp) +
  scale_y_continuous(labels = scales::percent) +
  labs(title    = "Figure 2 — On-Time vs Delayed by Tax Type",
       subtitle = "VAT: 100% on time | CIT: highest delay severity",
       x = "Tax Type", y = "Proportion of Filings", fill = "Status") +
  theme(legend.position = "top")
```

### Figure 3 — Filing Delay Distribution by Tax Type

```{r fig3, echo=FALSE, fig.height=4.5}
ggplot(df, aes(x = tax_type, y = filing_delay_days, fill = tax_type)) +
  geom_boxplot(outlier.colour = RED, outlier.size = 2.5,
               alpha = 0.75, width = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 23,
               size = 4, fill = "white") +
  scale_fill_manual(values = pal_tax) +
  scale_y_continuous(labels = scales::comma) +
  labs(title    = "Figure 3 — Filing Delay Distribution by Tax Type",
       subtitle = "Diamond = mean | Box = IQR | Red dots = outliers",
       x = "Tax Type", y = "Filing Delay (Days)") +
  guides(fill = "none")
```

### Figure 4 — Mean Filing Delay by Revenue Band & Management Type

```{r fig4, echo=FALSE, fig.height=5}
df |>
  dplyr::group_by(revenue_band, management_type) |>
  dplyr::summarise(mean_delay = mean(filing_delay_days),
                   .groups = "drop") |>
  ggplot(aes(x = revenue_band, y = mean_delay, fill = management_type)) +
  geom_col(position = "dodge", width = 0.6, colour = "white") +
  geom_text(aes(label = round(mean_delay, 1)),
            position = position_dodge(width = 0.6),
            vjust = -0.4, size = 3.2, colour = NAVY) +
  scale_fill_manual(values = pal_mgmt) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title    = "Figure 4 — Mean Filing Delay by Revenue Band & Management Type",
       subtitle = "Smaller firms record higher average delays",
       x = "Revenue Band", y = "Mean Delay (Days)", fill = "Management") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1),
        legend.position = "top")
```

### Figure 5 — Proportion Delayed by Revenue Band

```{r fig5, echo=FALSE, fig.height=4.5}
df |>
  dplyr::count(revenue_band, compliance_status) |>
  dplyr::group_by(revenue_band) |>
  dplyr::mutate(pct = n / sum(n)) |>
  dplyr::filter(compliance_status == "Delayed") |>
  ggplot(aes(x = revenue_band, y = pct, fill = revenue_band)) +
  geom_col(width = 0.6, colour = "white", alpha = 0.88) +
  geom_text(aes(label = scales::percent(pct, accuracy = 0.1)),
            vjust = -0.4, size = 3.5, fontface = "bold", colour = NAVY) +
  scale_fill_manual(values = pal_band) +
  scale_y_continuous(labels = scales::percent,
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title    = "Figure 5 — Proportion of Delayed Filings by Revenue Band",
       x = "Revenue Band", y = "Proportion Delayed") +
  guides(fill = "none")
```

---

## 7. Hypothesis Testing

> **Technique 3 of 5** — ANOVA, chi-squared, post-hoc tests

### 7.1 H1: Does Filing Delay Differ Across Tax Types?

**H₀:** Mean filing delay is the same across all four tax types.
**H₁:** At least one tax type has a significantly different mean filing delay.

```{r h1-anova}
anova_model <- aov(filing_delay_days ~ tax_type, data = df)
summary(anova_model)
```

```{r h1-kruskal}
kruskal.test(filing_delay_days ~ tax_type, data = df)
```

```{r h1-posthoc, echo=FALSE}
df |>
  rstatix::dunn_test(filing_delay_days ~ tax_type,
                     p.adjust.method = "BH") |>
  dplyr::select(group1, group2, statistic, p.adj, p.adj.signif) |>
  kbl(caption = "Table 7 — Dunn Post-Hoc Pairwise Comparisons",
      booktabs = TRUE, digits = 4) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover")) |>
  row_spec(0, bold = TRUE, background = NAVY, color = "white")
```

::: {.callout-tip title="Result — Hypothesis 1"}
H₀ is **rejected** (p < 0.01). Filing delay differs significantly across tax types.
Post-hoc tests show that VAT and WHT differ significantly from CIT, confirming the CIT
compliance gap is structural, not random.
:::

### 7.2 H2: Is Management Type Associated with Compliance Status?

**H₀:** No association between management type and compliance status.
**H₁:** Management type and compliance status are associated.

```{r h2-chi}
ct <- table(df$management_type, df$compliance_status)
print(ct)
chi_result <- chisq.test(ct)
print(chi_result)
cramers_v <- sqrt(chi_result$statistic / (sum(ct) * (min(dim(ct)) - 1)))
cat("\nCramer's V:", round(cramers_v, 4), "— negligible effect\n")
```

::: {.callout-tip title="Result — Hypothesis 2"}
The chi-squared test fails to reject H₀ (p > 0.05). Management type is **not**
significantly associated with compliance status. Tax type is a far stronger
differentiator than ownership structure.
:::

---

## 8. Correlation Analysis

> **Technique 4 of 5** — Pearson, Spearman; correlation vs causation

```{r corr, echo=FALSE, fig.height=5}
df_num <- df |>
  dplyr::select(filing_delay_days, filing_frequency, turnover_ngn, amount_paid_n) |>
  tidyr::drop_na()

corr_pearson <- cor(df_num, method = "pearson")

ggcorrplot(corr_pearson, method = "square", type = "lower",
           lab = TRUE, lab_size = 4,
           colors = c(RED, "white", NAVY),
           outline.col = "white", ggtheme = theme_lbs(),
           title = "Figure 6 — Pearson Correlation Heatmap")
```

```{r corr-spearman, echo=FALSE}
corr_sp <- cor(df_num, method = "spearman")

corr_sp |>
  as.data.frame() |>
  tibble::rownames_to_column("Predictor") |>
  dplyr::select(Predictor, filing_delay_days) |>
  dplyr::filter(Predictor != "filing_delay_days") |>
  dplyr::rename(`Spearman r` = filing_delay_days) |>
  dplyr::mutate(`Spearman r` = round(`Spearman r`, 4)) |>
  kbl(caption = "Table 8 — Spearman Correlations with Filing Delay",
      booktabs = TRUE) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover")) |>
  row_spec(0, bold = TRUE, background = NAVY, color = "white")
```

**Key observations:**
Filing frequency has the strongest correlation with delay (r ≈ −0.29) — monthly filers
outperform annual filers. Company turnover and amount paid show only weak relationships
with delay. Correlation does not imply causation; the relationships reflect structural
features of the tax system rather than direct causal mechanisms.

---

## 9. Regression Analysis

> **Technique 5 of 5** — OLS regression with diagnostics

```{r regression, echo=FALSE}
model_ols <- lm(filing_delay_days ~ tax_type + management_type +
                  industry_sector + turnover_ngn + amount_paid_n,
                data = df)

broom::tidy(model_ols, conf.int = TRUE) |>
  dplyr::mutate(across(where(is.numeric), ~ round(., 4)),
                significant = if_else(p.value < 0.05, "Yes", "No")) |>
  kbl(caption = "Table 9 — OLS Regression Coefficients", booktabs = TRUE) |>
  kable_styling(full_width = TRUE, bootstrap_options = c("striped","hover")) |>
  row_spec(0, bold = TRUE, background = NAVY, color = "white")
```

```{r reg-fit, echo=FALSE}
broom::glance(model_ols) |>
  dplyr::select(r.squared, adj.r.squared, sigma, statistic, p.value) |>
  dplyr::mutate(across(everything(), ~ round(., 4))) |>
  kbl(caption = "Table 10 — Model Fit Statistics", booktabs = TRUE) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover")) |>
  row_spec(0, bold = TRUE, background = NAVY, color = "white")
```

```{r diagnostics, echo=FALSE, fig.height=6}
par(mfrow = c(2, 2), bg = "#f7f9fc")
plot(model_ols, col = NAVY, pch = 20, cex = 0.8)
par(mfrow = c(1, 1))
```

```{r vif, echo=FALSE}
vif_vals <- car::vif(model_ols)
if (is.matrix(vif_vals)) {
  vif_df <- data.frame(Predictor = rownames(vif_vals), VIF = round(vif_vals[, 1], 3))
} else {
  vif_df <- data.frame(Predictor = names(vif_vals), VIF = round(vif_vals, 3))
}
vif_df |>
  kbl(caption = "Table 11 — Variance Inflation Factors", booktabs = TRUE) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover")) |>
  row_spec(0, bold = TRUE, background = NAVY, color = "white")
```

::: {.callout-tip title="Coefficient Interpretation"}
PAYE, VAT, and WHT each show significantly fewer delay days than the CIT reference
category, after controlling for company size and amount paid. The CIT compliance gap is
not explained by company size or remittance amount — it is an obligation-type effect.
The model R² indicates the proportion of variation in delay that is explained by these
predictors. The remaining variation reflects factors not captured here, such as internal
client capacity, FIRS portal availability, and management attention.
:::

---

## 10. Integrated Findings

::: {.callout-important title="Single Integrated Finding"}
Filing compliance failure in this dataset is a concentrated, obligation-type-specific
problem centred on CIT. It persists independently of company size, management structure,
and industry sector.
:::

The five techniques build the conclusion cumulatively:

- **EDA** reveals zero-inflation (75% on time) and extreme CIT outliers.
- **Visualisation** shows VAT at 100% on time; CIT with the highest severity.
- **Hypothesis Testing** confirms tax-type differences are statistically significant.
- **Correlation** identifies filing frequency as the strongest numerical correlate.
- **Regression** quantifies that PAYE, VAT, and WHT are each associated with substantially
  fewer delay days than CIT, while company size and amount paid are not significant.

**Recommendation:** Implement a tiered, obligation-specific monitoring protocol — placing
CIT on a dedicated 90/60/30-day pre-deadline alert schedule, introducing monthly estimated
CIT payments modelled on upstream oil and gas practice, and applying lighter-touch
monthly monitoring for VAT and PAYE.

---

## 11. Limitations & Further Work

::: {.callout-warning}
**Sample scope:** 104 observations across 13 clients within a single advisory portfolio,
not a random sample of Nigerian taxpayers.

**Cross-sectional structure:** Each client contributes multiple filings; standard errors
may be understated. A mixed-effects model would be more appropriate.

**Zero-inflation:** Severe zero-inflation violates OLS assumptions. A hurdle model
combining logistic regression with truncated regression would be more suitable.

**Filing_Frequency collinearity:** Perfectly collinear with Tax_Type (CIT is the only
annual filer) and was not included as an independent predictor.

**With more data, time, or computing power:** extend the panel to additional years for
fixed-effects regression; apply a zero-inflated negative binomial model; incorporate FIRS
penalty data; build a machine-learning classifier to predict next-period non-compliance.
:::

---

## References

Adi, B. (2026). *AI-powered business analytics: A practical textbook for data-driven
decision making.* Lagos Business School / markanalytics.online. https://markanalytics.online

R Core Team. (2024). *R: A language and environment for statistical computing.* R
Foundation for Statistical Computing. https://www.R-project.org/

Wickham, H., et al. (2019). Welcome to the tidyverse. *Journal of Open Source Software,
4*(43), 1686. https://doi.org/10.21105/joss.01686

Wickham, H. (2016). *ggplot2: Elegant graphics for data analysis.* Springer.

Obadan, L. (2026). *Corporate tax compliance filing records — anonymised client dataset*
[Dataset]. Compiled from operational compliance tracking registers, Lagos, Nigeria,
January 2024 to July 2025. Data available on request from the author.

---

## Appendix: AI Usage Statement

Claude (Anthropic) was used to assist with structuring the Quarto document, generating
initial R code scaffolding, and suggesting appropriate diagnostic approaches. All
analytical decisions — selection of the outcome variable, choice of statistical tests,
interpretation of coefficients, and the integrated recommendation — were made
independently by the author. The dataset was compiled from records within my own
professional practice, and no data was shared with any AI tool at any stage.
